A semantics of core AUTOSAR

### 2016-04-27 paper/paper.tex
LaTeX translation of semantics done.
Edits during PaJa + JoNo meeting
Paper reading + updates pages 1--3

Unfortunately, the AUTOSAR specification is not very rigorous, despite a sheer size of more than one hundred pdf documents and over 12,500 pages of text and UML diagrams in total. It is also not very abstract, in that it makes frequent references to assumed implementation techniques for the purpose of defining its semantics. In practice, the AUTOSAR standard becomes blurred with the specific behavior of one's chosen platform and development tools. And because the standard is open to interpretation, the interoperability of software components across tools and platforms is often seriously hampered. What is more, a single software component cannot easily be studied and understood in isolation, since its interactive behavior is only indirectly defined in terms of the concrete C-code and OS tasks that realize it and interaction environment.

This paper takes an important step towards a remedy to these problems, by contributing a formal specification of a substantial core of the AUTOSAR standard. The formalization covers most of the Software Component Template and its accompanying Run-Time Environment (RTE) (Section~\ref{sec:Calc}), and is able to directly express every legal way a system of SWCs can evolve in at run-time on an arbitrarily fast platform (Section~\ref{sec:Sem}). It can thus serve as a basis for both a concrete AUTOSAR implementation on a specific platform (whose supported behaviors must be a subset of those defined by the formal semantics), and a platform-independent AUTOSAR simulator (where behaviors from the legal set can be picked at random). Examples of semantic derivations and their applications are found in Section~\ref{sec:Examples}.

@@

A runnable under execution is called a \textbf{runnable instance}, and these may execute concurrently even if they belong to the same SWC. Each runnable indicates whether it may be instantiated concurrently with itself. Runnable code can also use \textbf{exclusive areas} (a form of mutex semaphores) to further control concurrent behavior.

To interact with ports, inter-runnable variables and exclusive areas, runnables utilize what is abstractly known as the \textbf{virtual function bus} (VFB). Concretely, the VFB takes the shape of a C programming interface to the \textbf{run-time environment} (RTE), custom-generated for each runnable on basis of its access constraints. Behind the RTE, a collection of OS kernels, communication stacks and other basic software modules implement the VBF services for the platform at hand. Complete AUTOSAR designs also typically contain \textbf{ECU} and \textbf{task assignments}, which are manually built tables that control how SWCs and runnables map onto the available hardware and OS resources, respectively.

AUTOSAR models are completely static, which means that all components, runnables, ports, connectors, etc, are created ahead of execution time. Runnable instances are an exception, but they are only identified by AUTOSAR for conceptual purposes and never part of any model syntax. This static nature renders AUTOSAR models very suitable to graphical notations, which by far is the most widespread representation format for AUTOSAR models. An graphical model example is shown in Figure~\ref{fig:model}.

@@

Components and ports do not carry any dynamic state besides the state of their constituent parts, so they need no explicit representation as process terms. However, in order to still be able to talk about model elements using their locally scoped names, we employ a hierarchical addressing scheme, such that address $\adr{R:I}$ means runnable $R$ of component $I$ and $\adr{E:P:I}$ means element $E$ of port $P$ within component $I$. The following meta-variable conventions will be used:
%
\[
\begin{array}{rcll}
	I	& \in		& \text{Component names} \\
	R	& \in		& \text{Runnable names} \\
	U	& \in		& \text{Port names} \\
	E	& \in		& \text{Sender-receiver element names} \\
	O	& \in		& \text{Client-server operation names} \\
	S	& \in		& \text{Inter-runnable variable names} \\
	X	& \in		& \text{Exclusive area names} \\
	\\
	A	& \in		& \text{Addresses} \\
	A	& ::=		& \adr{R:I}   \sep   \adr{E:P:I}   \sep   \adr{O:U:I}   \sep   \adr{S:I}   \sep   \adr{X:I}   \\
\end{array}
\]
%
Since AUTOSAR components can be arbitrarily nested, the component names $I$ can also be considered to have a similarly nested internal structure (component $I_1$ within component $I_2$ within component $I_3$, etc). We can fully ignore this detail here, though, as the hierarchical layout of components in an AUTOSAR system has no run-time implications \cite{TODO2}.

The process terms of our calculus are as follows:
\begin{itemize}
\item $\runnable{\adr{R:I}}{T}{\Act}{N}$  \newline
Common state for runnable $R$ of component $I$, showing $N$ current instances, $T$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\rinst{\adr{R:I}}{C}{\Xs}{\Code}$ \newline
An instance of runnable $R$ of component $I$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $C$.
\item $\excl{\adr{X:I}}{B}$ \newline
Exclusive area $X$ of component $I$, with boolean occupation state $B$.
\item $\irv{\adr{S:I}}{V}$ \newline
Inter-runnable variable $S$ of component $I$, currently holding value $V$.
\item $\delem{\adr{E:U:I}}{B}{V}$ \newline
Non-queued sender-receiver data element $E$ in port $U$ of component $I$, currently holding value $V$ with boolean update status flag $B$.
\item $\qelem{\adr{E:U:I}}{\Vs}{TODO}$ \newline
Queued sender-receiver data element $E$ in port $U$ of component $I$, currently holding value sequence $\Vs$.
\item $\opres{\adr{O:U:I}}{\Vs}$ \newline
Client-server operation $O$ in port $U$ of component $I$, currently holding result value sequence $\Vs$.
\item $\timer{\adr{R:I}}{T_p}{T}$ \newline  %TODO: timer has only 2 args in the semantics
Timer state for runnable $R$ of component $I$ and period $T_p$, with $T$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $V$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\void$ and the error codes that might be returned from RTE calls. Meta-variables $N$, $B$ and $T$ range over natural numbers, boolean values and (floating point) time values, respectively. The client $C$ of a runnable instance is the address of the invoking port operation (if the runnable was invoked by a client-server call), or $\void$ otherwise. An activation state $\Act$ either toggles between $\idle$ and $\pending$, or it has the special shape $\serving{\Cs}{\Vs}$ (in the case of a runnable triggered by client-server calls). Formally,
$$
\begin{array}{rcll}
	V	& \in		& \text{Values} \\
	N	& \in		& \text{Natural numbers} \\
	B	& \in		& \text{Boolean values} \\
	T	& \in		& \text{Time values} \\
	C	& ::=		& \adr{O:U:I} \sep \void \\
	\Act	& ::=		& \idle \sep \pending \sep \serving{\Cs}{\Vs} \\
\end{array}
$$
We write $\Vs$ for a sequence of values, $\Cs$ for a sequence of clients, etc. Sequences may be empty, which is written $\epsilon$. Left (right) concatenation of an element with a sequence is written $\adr{V:Vs}$ ($\adr{Vs:V}$).

@@

Together with an operator for composing process terms in parallel, the process grammar for our AUTOSAR calculus becomes
$$
\begin{array}{rcll}
	P,Q 	& ::= &
		\runnable{A}{K}{T}{\Act}{N}	\\
		& | &
		\rinst{A}{C}{\Xs}{\Code}	\\
		& | &
		\excl{A}{B}			\\
		& | &
		\irv{A}{V}				\\
		& | &
		\delem{A}{B}{V}			\\
		& | &
		\qelem{A}{\Vs}{TODO}			\\
		& | &
		\opres{A}{\Vs}			\\
		& | &
		\timer{A}{T_p}{T}		\\
		& | &
		P|Q					\\
\end{array}
$$

@@

The label $L$ of a transition can be ow two kinds, indicating that the affected process term either "says" or "hears" something during the transition. A label also always consists of an address $A$ paired with some additional detail $D$, as captured in the following label grammar:
\[
\begin{array}{rcll}
	L	& ::=		& A!D		& \hspace{4em}\text{Say $A$ and $D$} \\
		&		& A?D		& \hspace{4em}\text{Hear $A$ and $D$} \\
\end{array}
\]
The possible forms of the extra label information $D$ will be introduced as the different transition rules are defined.

@@

\item For each runnable $R$ of each component $I$, a term
$$
\begin{array}{c}
	\runnable{\adr{R:I}}{TODO}{0.0}{\Act}{0}
\end{array}
$$
where $\Act$ is $\serving{\epsilon}{\epsilon}$ if $R$ is triggered by an \emph{OperationInvokedEvent}; else $\Act$ is $\pending$ if $R$ is triggered by an \emph{InitEvent}; otherwise $\Act$ is $\idle$.
% SWS_Rte_03524 (SWS_Rte_03526?) forbids mixing OperationInvokedEvents with other events for the same runnable

\item For each exclusive area $X$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\excl{\adr{X:I}}{\false}
\end{array}
$$

\item For each inter-runnable variable $S$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\irv{\adr{S:I}}{V}
\end{array}
$$
where $V$ is the initial value attributed to $S$ if it exists, otherwise $V$ is $\void$.

\item For each data element $E$ of each required port $U$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\qelem{\adr{E:U:I}}{\epsilon}{TODO}
\end{array}
$$
if the \emph{swImplPolicy} attribute of element $E$ is \emph{queued}; otherwise, a term
$$
\begin{array}{c}
	\delem{\adr{E:U:I}}{\false}{\text{NO\_DATA}}
\end{array}
$$

\item For each operation $O$ of each required port $U$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\opres{\adr{O:U:I}}{\epsilon}
\end{array}
$$

\item For each \emph{TimingEvent} triggering each runnable $R$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\timer{\adr{R:I}}{T}{0.0}
\end{array}
$$
where $T$ is the period of the event.

@@

Some further static model information will also be referenced by the transition rules. We refer to the \emph{connectors} of a model using a binary relation $\Rightarrow$ on addresses, defined as
$$
	\Tconnect{\adr{E:$U_1$:$I_1$}}{\adr{E:$U_2$:$I_2$}} \;\;\text{for all elements $E$ of port $U_2$}
$$
if there is a connector in the model from provided sender-receiver port $U_1$ of component $I_1$ to required port $U_2$ of component $I_2$; and
$$
	\Tconnect{\adr{O:$U_1$:$I_1$}}{\adr{O:$U_2$:$I_2$}} \;\;\text{for all operations $O$ of port $U_1$}
$$
if there is a connector in the model from required client-server port $U_1$ of component $I_1$ to provided port $U_2$ of component $I_2$.

@@

\subsection{Exclusive areas}

@@

% \inputminted{prolog}{../prolog/semantics.pl}

\input{../prolog/tool/semantics.tex}

% \verbatiminput{../prolog/semantics.pl}

>>>>

Unfortunately, the AUTOSAR specification is not very rigorous, despite a sheer size of more than one hundred pdf documents and over 12,500 pages of text and UML diagrams in total. It is also not very abstract, in that it makes frequent references to assumed implementation techniques for the purpose of defining its semantics. In practice, the AUTOSAR standard becomes blurred with the specific behavior of one's chosen platform and development tools. And because the standard is open to interpretation, the interoperability of software components across tools and platforms is often seriously hampered. What is more, a single software component cannot easily be studied and understood in isolation, since its interactive behavior is only indirectly defined in terms of the concrete C-code and OS tasks that realize it and its interaction environment.

This paper takes an important step towards a remedy to these problems, by contributing a formal specification of a substantial core of the AUTOSAR standard. The formalization covers most of the Software Component Template and its accompanying Run-Time Environment (RTE) (Section~\ref{sec:Calc}), and is able to directly express every legal way a system of SWCs can evolve at run-time on an arbitrarily fast platform (Section~\ref{sec:Sem}). It can thus serve as a basis for both a concrete AUTOSAR implementation on a specific platform (whose supported behaviors must be a subset of those defined by the formal semantics), and a platform-independent AUTOSAR simulator (where behaviors from the legal set can be picked at random). Examples of semantic derivations and their applications are found in Section~\ref{sec:Examples}.

@@

A runnable under execution is called a \textbf{runnable instance}, and these may execute concurrently even if they belong to the same SWC. Each runnable indicates whether or not it may be instantiated concurrently with itself. Runnable code can also use \textbf{exclusive areas} (a form of mutex semaphores) to further control concurrent behavior.

To interact with ports, inter-runnable variables and exclusive areas, runnables use what is abstractly known as the \textbf{virtual function bus} (VFB). Concretely, the VFB takes the shape of a C programming interface to the \textbf{run-time environment} (RTE), custom-generated for each runnable on the basis of its access constraints. Behind the RTE, a collection of OS kernels, communication stacks and other basic software modules implement the VFB services for the platform at hand. Complete AUTOSAR designs also typically contain \textbf{ECU} and \textbf{task assignments}, which are manually built tables that control how SWCs and runnables map onto the available hardware and OS resources, respectively.

AUTOSAR models are completely static, which means that all components, runnables, ports, connectors, etc, are created ahead of execution time. Runnable instances are an exception, but they are only identified by AUTOSAR for conceptual purposes and never part of any model syntax. This static nature renders AUTOSAR models very suitable to graphical notations, which by far is the most widespread representation format for AUTOSAR models. A graphical model example is shown in Figure~\ref{fig:model}.

TODO: add fig:model

@@

Components and ports do not carry any dynamic state besides the state of their constituent parts, so they need no explicit representation as process terms. However, in order to still be able to talk about model elements using their locally scoped names, we employ a hierarchical addressing scheme, such that address $\adrr{R}{I}$ means runnable $R$ of component $I$ and $\adrep{E}{P}{I}$ means element $E$ of port $P$ within component $I$. We use the following meta-variable conventions:

%TODO: port name convention change from U to P? (see ../prolog/semantics.eprover)
%
\[
\begin{array}{rcll}
	I	& \in		& \text{Component names} \\
	R	& \in		& \text{Runnable names} \\
	U	& \in		& \text{Port names} \\
	E	& \in		& \text{Sender-receiver element names} \\
	O	& \in		& \text{Client-server operation names} \\
	S	& \in		& \text{Inter-runnable variable names} \\
	X	& \in		& \text{Exclusive area names} \\
	\\
	A	& \in		& \text{Addresses} \\
	A	& ::=		& \adrr{R}{I}   \sep   \adrep{E}{P}{I}   \sep \adrop{O}{P}{I}   \sep   \adrs{S}{I}   \sep   \adrx{X}{I}   \\
\end{array}
\]
%
Since AUTOSAR components can be arbitrarily nested, the component names $I$ can also be considered to have a similarly nested internal structure (component $I_1$ within component $I_2$ within component $I_3$, etc). We can fully ignore this detail here, though, as the hierarchical layout of components in an AUTOSAR system has no run-time implications \cite{TODO2}.

The process terms of our calculus are as follows:
\begin{itemize}
\item $\runnable{\adrr{R}{I}}{TODOK}{T}{\Act}{N}$  \newline
Common state for runnable $R$ of component $I$, showing $N$ current instances, $T$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\rinst{\adrr{R}{I}}{C}{\Xs}{\Code}$ \newline
An instance of runnable $R$ of component $I$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $C$.
\item $\excl{\adrx{X}{I}}{B}$ \newline
Exclusive area $X$ of component $I$, with boolean occupation state $B$.
\item $\irv{\adrs{S}{I}}{V}$ \newline
Inter-runnable variable $S$ of component $I$, currently holding value $V$.
\item $\delem{\adrep{E}{U}{I}}{B}{V}$ \newline
Non-queued sender-receiver data element $E$ in port $U$ of component $I$, currently holding value $V$ with boolean update status flag $B$.
\item $\qelem{\adrep{E}{U}{I}}{\Vs}{TODO}$ \newline
Queued sender-receiver data element $E$ in port $U$ of component $I$, currently holding value sequence $\Vs$.
\item $\opres{\adrop{O}{U}{I}}{\Vs}$ \newline
Client-server operation $O$ in port $U$ of component $I$, currently holding result value sequence $\Vs$.
\item $\timer{\adrr{R}{I}}{T}$ \newline  %TODO: timer has only 2 args in the semantics so {T_p} is not described
Countdown timer state for runnable $R$ of component $I$ with $T$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $V$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\void$ and the error codes that might be returned from RTE calls. Meta-variables $N$, $B$ and $T$ range over natural numbers, boolean values and (floating point) time values, respectively. The client $C$ of a runnable instance is the address of the invoking port operation (if the runnable was invoked by a client-server call), or $\void$ otherwise. An activation state $\Act$ either toggles between $\idle$ and $\pending$, or it has the special shape $\serving{\Cs}{\Vs}$ (in the case of a runnable triggered by client-server calls). Formally,
$$
\begin{array}{rcll}
	V	& \in		& \text{Values} \\
	N	& \in		& \text{Natural numbers} \\
	B	& \in		& \text{Boolean values} \\
	T	& \in		& \text{Time values} \\
	C	& ::=		& \adrop{O}{U}{I} \sep \void \\
	\Act	& ::=		& \idle \sep \pending \sep \serving{\Cs}{\Vs} \\
\end{array}
$$
We write $\Vs$ for a sequence of values, $\Cs$ for a sequence of clients, etc. Sequences may be empty, which is written $\epsilon$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

@@

Together with an operator for composing process terms in parallel, the process grammar for our AUTOSAR calculus becomes
$$
\begin{array}{rcll}
	P,Q 	& ::= &
		\runnable{A}{K}{T}{\Act}{N}	\\
		& | &
		\rinst{A}{C}{\Xs}{\Code}	\\
		& | &
		\excl{A}{B}			\\
		& | &
		\irv{A}{V}				\\
		& | &
		\delem{A}{B}{V}			\\
		& | &
		\qelem{A}{\Vs}{TODO}			\\
		& | &
		\opres{A}{\Vs}			\\
		& | &
		\timer{A}{T}		\\
		& | &
		P|Q					\\
\end{array}
$$

@@

The label $L$ of a transition can be of two kinds, indicating that the affected process term either "says" or "hears" something during the transition. A label also always consists of an address $A$ paired with some additional detail $D$, as captured in the following label grammar:
\[
\begin{array}{rcll}
	L	& ::=		& A!D		& \hspace{4em}\text{Say $A$ and $D$} \\
		&	|	& A?D		& \hspace{4em}\text{Hear $A$ and $D$} \\
\end{array}
\]
The possible forms of the extra label information $D$ will be introduced as the different transition rules are defined.

@@

\item For each runnable $R$ of each component $I$, a term
$$
\begin{array}{c}
	\runnable{\adrr{R}{I}}{TODO}{0.0}{\Act}{0}
\end{array}
$$
where $\Act$ is $\serving{\epsilon}{\epsilon}$ if $R$ is triggered by an \emph{OperationInvokedEvent}; else $\Act$ is $\pending$ if $R$ is triggered by an \emph{InitEvent}; otherwise $\Act$ is $\idle$.
% SWS_Rte_03524 (SWS_Rte_03526?) forbids mixing OperationInvokedEvents with other events for the same runnable

\item For each exclusive area $X$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\excl{\adrx{X}{I}}{\false}
\end{array}
$$

\item For each inter-runnable variable $S$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\irv{\adrs{S}{I}}{V}
\end{array}
$$
where $V$ is the initial value attributed to $S$ if it exists, otherwise $V$ is $\void$.

\item For each data element $E$ of each required port $U$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\qelem{\adrep{E}{U}{I}}{\epsilon}{TODO}
\end{array}
$$
if the \emph{swImplPolicy} attribute of element $E$ is \emph{queued}; otherwise, a term
$$
\begin{array}{c}
	\delem{\adrep{E}{U}{I}}{\false}{\text{nodata}}
\end{array}
$$

\item For each operation $O$ of each required port $U$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\opres{\adrep{O}{U}{I}}{\epsilon}
\end{array}
$$

\item For each \emph{TimingEvent} triggering each runnable $R$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\timer{\adrr{R}{I}}{0.0}
\end{array}
$$
% where $T$ is the period of the event.
%TODO: Does this mean all timed runnables will want to act _immediately_ (at the same time)? No init delay?

@@

Some further static model information will also be referenced by the transition rules. We refer to the \emph{connectors} of a model using a binary relation $\Rightarrow$ on addresses, defined as
$$
	\Tconnect{\adrep{E}{$U_1$}{$I_1$}}{\adrep{E}{$U_2$}{$I_2$}} \;\;\text{for all elements $E$ of port $U_2$}
$$
if there is a connector in the model from provided sender-receiver port $U_1$ of component $I_1$ to required port $U_2$ of component $I_2$; and
$$
	\Tconnect{\adrop{O}{$U_1$}{$I_1$}}{\adrop{O}{$U_2$}{$I_2$}} \;\;\text{for all operations $O$ of port $U_1$}
$$
if there is a connector in the model from required client-server port $U_1$ of component $I_1$ to provided port $U_2$ of component $I_2$.

@@

%TODO: Patrik got here 2016-04-27 in a reading + macro conversion round
\subsection{Exclusive areas}

@@

\input{semantics.tex}
%\input{../prolog/tool/semantics.tex}
